quollr: An R Package for Visualizing 2-D Models from Nonlinear Dimension Reductions in High-Dimensional Space

### 2025-12-01 paper-quollr.Rmd
Added package installation code chunk

---
title: "quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space"
abstract: >
  Non-linear dimension reduction (NLDR) methods provide a low-dimensional representation of high-dimensional data ($p\text{-}D$) by applying a non-linear transformation. However, the complexity of the transformations and data structures can create wildly different representations depending on the method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures. The R package \CRANpkg{quollr} has been developed as a new visual tool to determine which method and which hyper-parameter choices provide the most accurate representation of high-dimensional data. The `scurve` data from the package is used to illustrate the algorithm. Single-cell RNA sequencing (scRNA-seq) data from mouse limb muscles are used to demonstrate the usability of the package.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-quollr.bib
editor_options: 
  chunk_output_type: inline
---

@@

options(repos = c(CRAN = "https://cran.rstudio.com")) # Setup mirror

packages_to_check <- c("remotes", "quollr", "tibble", "knitr", "kableExtra", "ggplot2", "dplyr", "patchwork", "readr", "plotly", "crosstalk", "htmltools")

for (pkg in packages_to_check) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    install.packages(pkg)
  } else {
    installed_version <- packageVersion(pkg)
    available_version <- tryCatch({
      utils::packageDescription(pkg)$Version
    }, error = function(e) NA) # Handle cases where package info isn't readily available

    if (!is.na(available_version) && installed_version < package_version(available_version)) {
      message(paste("A newer version of package", pkg, "is available. Updating..."))
      install.packages(pkg)
    } else {
      message(paste("Package", pkg, "is up-to-date (version", installed_version, ")."))
    }
  }
}

@@

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021], can create hugely different representations depending on the selected method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

```{r, echo=TRUE, eval=TRUE}
model_highd <- fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
```

@@

```{r, echo=TRUE, eval=TRUE}
combined_data <- comb_data_model(
  highd_data = scurve,
  model_highd = model_highd$model_highd,
  model_2d = model_highd$model_2d
)

tour_view <- show_langevitour(
  point_data = combined_data,
  edge_data = model_highd$trimesh_data
)
```

@@

The result is a `tibble` of the scaled $2\text{-}D$ embedding with an additional `h` column, indicating the hexagonal bin to which each point belongs.

>>>>

---
title: "quollr: An R Package for Visualizing 2-D Models from Nonlinear Dimension Reductions in High-Dimensional Space"
abstract: >
  Nonlinear dimension reduction methods provide a low-dimensional representation of high-dimensional data by applying a Nonlinear transformation. However, the complexity of the transformations and data structures can create wildly different representations depending on the method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures. The R package quollr has been developed as a new visual tool to determine which method and which hyper-parameter choices provide the most accurate representation of high-dimensional data. The scurve data from the package is used to illustrate the algorithm. Single-cell RNA sequencing (scRNA-seq) data from mouse limb muscles are used to demonstrate the usability of the package.
draft: true
author:  
  - name: Jayani P. Gamage
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{float}
output: 
 rjtools::rjournal_article:
    keep_md: true
bibliography: paper-quollr.bib
editor_options: 
  chunk_output_type: inline
---

@@

# Ensure remotes is available
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# Read the package list
pkgs_raw <- readLines("_Rpackages.txt")
pkgs_raw <- pkgs_raw[nzchar(pkgs_raw)]  # remove empty lines

# Split into package and version
parts <- strsplit(pkgs_raw, "==")
pkgs <- vapply(parts, `[[`, "", 1)
versions <- vapply(parts, `[[`, "", 2)

# Install packages with exact versions
for (i in seq_along(pkgs)) {
  pkg <- pkgs[i]
  ver <- versions[i]

  if (!requireNamespace(pkg, quietly = TRUE) ||
      as.character(packageVersion(pkg)) != ver) {

    message(sprintf("Installing %s (%s)", pkg, ver))
    remotes::install_version(pkg, version = ver, upgrade = "never")
  }
}

@@

Nonlinear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2019], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021], can create hugely different representations depending on the selected method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

```{r, echo=TRUE, eval=TRUE}
model_obj <- fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
```

@@

```{r, echo=TRUE, eval=TRUE}
combined_data <- comb_data_model(
  highd_data = scurve,
  model_highd = model_obj$model_highd,
  model_2d = model_obj$model_2d
)

tour_view <- show_langevitour(
  point_data = combined_data,
  edge_data = model_obj$trimesh_data
)
```

@@

The result is a `tibble` of the scaled $2\text{-}D$ embedding with an additional `h` column, indicating the hexagonal bin to which each point belongs. 
  ?

### 2025-12-01 paper-quollr.tex
Added package installation code chunk

% !TeX root = RJwrapper.tex
\title{quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space}

@@

\abstract{%
Non-linear dimension reduction (NLDR) methods provide a low-dimensional representation of high-dimensional data (\(p\text{-}D\)) by applying a non-linear transformation. However, the complexity of the transformations and data structures can create wildly different representations depending on the method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures. The R package \CRANpkg{quollr} has been developed as a new visual tool to determine which method and which hyper-parameter choices provide the most accurate representation of high-dimensional data. The \texttt{scurve} data from the package is used to illustrate the algorithm. Single-cell RNA sequencing (scRNA-seq) data from mouse limb muscles are used to demonstrate the usability of the package.
}

@@

Non-linear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and projection (UMAP) \citep{leland2018}, potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019}, large-scale dimensionality reduction Using triplets (TriMAP) \citep{amid2019}, and pairwise controlled manifold approximation (PaCMAP) \citep{yingfan2021}, can create hugely different representations depending on the selected method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

\begin{verbatim}
model_highd <- fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
\end{verbatim}

@@

\begin{verbatim}
combined_data <- comb_data_model(
  highd_data = scurve,
  model_highd = model_highd$model_highd,
  model_2d = model_highd$model_2d
)

tour_view <- show_langevitour(
  point_data = combined_data,
  edge_data = model_highd$trimesh_data
)
\end{verbatim}

@@

The result is a \texttt{tibble} of the scaled \(2\text{-}D\) embedding with an additional \texttt{h} column, indicating the hexagonal bin to which each point belongs.

>>>>

% !TeX root = RJwrapper.tex
\title{quollr: An R Package for Visualizing 2-D Models from Nonlinear Dimension Reductions in High-Dimensional Space}

@@

\abstract{%
Nonlinear dimension reduction methods provide a low-dimensional representation of high-dimensional data by applying a Nonlinear transformation. However, the complexity of the transformations and data structures can create wildly different representations depending on the method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures. The R package quollr has been developed as a new visual tool to determine which method and which hyper-parameter choices provide the most accurate representation of high-dimensional data. The scurve data from the package is used to illustrate the algorithm. Single-cell RNA sequencing (scRNA-seq) data from mouse limb muscles are used to demonstrate the usability of the package.
}

@@

Nonlinear dimension reduction (NLDR) techniques, such as t-distributed stochastic neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and projection (UMAP) \citep{leland2018}, potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019}, large-scale dimensionality reduction Using triplets (TriMAP) \citep{amid2019}, and pairwise controlled manifold approximation (PaCMAP) \citep{yingfan2021}, can create hugely different representations depending on the selected method and hyper-parameter choices. It is difficult to determine whether any of these representations are accurate, which one is the best, or whether they have missed important structures.

@@

\begin{verbatim}
model_obj <- fit_highd_model(
  highd_data = scurve, 
  nldr_data = scurve_umap, 
  b1 = 21, 
  q = 0.1, 
  hd_thresh = 0)
\end{verbatim}

@@

\begin{verbatim}
combined_data <- comb_data_model(
  highd_data = scurve,
  model_highd = model_obj$model_highd,
  model_2d = model_obj$model_2d
)

tour_view <- show_langevitour(
  point_data = combined_data,
  edge_data = model_obj$trimesh_data
)
\end{verbatim}

@@

The result is a \texttt{tibble} of the scaled \(2\text{-}D\) embedding with an additional \texttt{h} column, indicating the hexagonal bin to which each point belongs.
?
